\textbf{Efficient Implementation Strategies for Block Ciphers on ARMv8}\\ {\footnotesize Bachelorarbeit}

### 2023-04-11 Paper/chapters/introduction.tex
rearrange table captions
more intro

The encryption process has an 18-round Feistel structure for 128-bit keys and a
24-round Feistel structure for 192/256-bit key and employs key whitening to
increase security. First, subkeys $kw_{t(64)}(t=0,1,2,3)$,
$k_{u(64)}(u=1,2,\dots,(17|23)$ and $kl_{v(64)}(v=0,1,2,3)$ are generated from
the master key. Then, pre-whitening keys are applied to the left and right half
of the plaintext $m_{(128)}=L_{(64)}||R_{(64)}$:

\[
    L\leftarrow L\oplus kw_0, R\leftarrow R\oplus kw_1
\]

The next steps differ for 128-bit and 192/256-bit keys:

\begin{table}[h!]
    \centering
    \caption{Camellia encryption}
    \begin{tabular}{lll}
        \toprule
        Round $r$ & 128-bit keys & 192/256-bit keys \\
        \midrule
        0-4 & $FE(L,R,k_r)$ & $FE(L,R,k_r)$ \\
        \midrule
        5 & {$\!\begin{aligned}&FE(L,R)\\&FLA(L,R,kl_0,kl_1)\end{aligned}$} & {$\!\begin{aligned}&FE(L,R)\\&FLA(L,R,kl_0,kl_1)\end{aligned}$} \\
        \midrule
        6-10 & $FE(L,R,k_r)$ & $FE(L,R,k_r)$ \\
        \midrule
        11 & {$\!\begin{aligned}&FE(L,R)\\&FLA(L,R,kl_2,kl_3)\end{aligned}$} & {$\!\begin{aligned}&FE(L,R)\\&FLA(L,R,kl_2,kl_3)\end{aligned}$} \\
        \midrule
        12-16 & $FE(L,R,k_r)$ & $FE(L,R,k_r)$ \\
        \midrule
        17 & & {$\!\begin{aligned}&FE(L,R)\\&FLA(L,R,kl_4,kl_5)\end{aligned}$} \\
        \midrule
        18-23 & & $FE(L,R,k_r)$ \\
        \bottomrule
    \end{tabular}
\end{table}

@@

\[
    c=(R\oplus kw_2)||(L\oplus kw_3)
\]

@@

$FE$ and $FLA$ are macros for applying one round of the Feistel structure and
$FL$ functions and are defined as follows:

@@

\begin{table}[h!]
    \centering
    \begin{tabularx}{\textwidth}{lX}
        \toprule
        Profile & Description \\
        \midrule
        Application (A) & Traditional use with virtual memory and privilege level support \\
        Real-time (R) & Real-time, low-latency, deterministic embedded systems \\
        Microcontroller (M) & Very low-power, fast-interrupt embedded systems \\
        \bottomrule
    \end{tabularx}
    \caption{ARMv8 profiles}
\end{table}

\begin{table}[h!]
    \centering
    \begin{tabularx}{\textwidth}{llX}
        \toprule
        Execution state & Usage & Instruction sets \\
        \midrule
        AArch32 & 32-bit compatibility & A32/T32 \\
        AArch64 & 64-bit & A64 \\
        \bottomrule
    \end{tabularx}
    \caption{ARMv8 execution states}
\end{table}

@@

\begin{table}[h!]
    \centering
    \small
    \begin{tabularx}{\textwidth}{llX}
        \toprule
        Addressing mode & Example & Description \\
        \midrule
        Base register & \texttt{LDR W0, [X1]} & \texttt{W0 = *(X1);} \\
        Offset & \texttt{LDR W0, [X1, \#12]} & \texttt{W0 = *(X1 + 12);} \\
        Pre-indexing & \texttt{LDR W0, [X1, \#12]!} & \texttt{X1 += 12; W0 = *(X1);} \\
        Post-indexing & \texttt{LDR W0, [X1], \#12} & \texttt{W0 = *(X1); X1 += 12;} \\
        \bottomrule
    \end{tabularx}
    \caption{A64 addressing modes}
\end{table}

@@

\begin{table}[h!]
    \centering
    \footnotesize
    \begin{tabularx}{\textwidth}{llX}
        \toprule
        Intrinsic && Summary \\
        \midrule
        \texttt{uint64\_t} & \texttt{vgetq\_lane\_u64(void)} & Extract a single lane \\
        \midrule
        \texttt{void} & \texttt{vsetq\_lane\_u64(uint64\_t)} & Insert a single lane \\
        \midrule
        \texttt{uint64x2\_t} & \texttt{vdupq\_n\_u64(uint64\_t)} & Initialize all lanes to same value \\
        \midrule
        \texttt{void} & \texttt{vst1q\_u64(uint64\_t*, uint64x2\_t)} & Store from register to memory \\
        \midrule
        \texttt{uint64x2\_t} & \texttt{vld1q\_u64(uint64\_t*, uint64x2\_t)} & Load from memory to register \\
        \midrule
        \texttt{uint8x16\_t} & \texttt{veorq\_u8(uint8x16\_t, uint8x16\_t)} & bitwise XOR \\
        \midrule
        \texttt{uint8x16\_t} & \texttt{vandq\_u8(uint8x16\_t, uint8x16\_t)} & bitwise AND \\
        \midrule
        \texttt{uint8x16\_t} & \texttt{vorrq\_u8(uint8x16\_t, uint8x16\_t)} & bitwise OR \\
        \midrule
        \texttt{uint8x16\_t} & \texttt{vmvnq\_u8(uint8x16\_t)} & bitwise NOT \\
        \midrule
        \texttt{uint8x16\_t} & \texttt{vqtbl2q\_u8(uint8x16\_t, uint8x16\_t)} & permutation (\texttt{TBL}) \\
        \bottomrule
    \end{tabularx}
    \caption{Common NEON intrinsics}
\end{table}

>>>>

The encryption process has an 18-round Feistel structure for 128-bit keys and a
24-round Feistel structure for 192/256-bit key and employs key whitening to
increase security. First, subkeys $kw_{t(64)}(t=0,1,2,3)$,
$k_{u(64)}(u=0,1,\dots,(17|23)$ and $kl_{v(64)}(v=0,1,2,3)$ are generated from
the master key. Then, pre-whitening keys are applied to the plaintext
$m_{(128)}=L_{(64)}||R_{(64)}$:

\[
    (L||R)\leftarrow (L||R)\oplus (kw_0||kw_1)
\]

The next steps differ for 128-bit and 192/256-bit keys in the number of rounds:

\begin{table}[h!]
    \centering
    \caption{Camellia encryption}
    \begin{tabular}{lll}
        \toprule
        Round $r$ & 128-bit keys & 192/256-bit keys \\
        \midrule
        0-4 & $(L||R)\leftarrow FE(L,R,k_r)$ & $(L||R)\leftarrow FE(L,R,k_r)$ \\
        \midrule
        5 & {$\!\begin{aligned}&(L||R)\leftarrow FE(L,R)\\&(L||R)\leftarrow FLL(L,R,kl_0,kl_1)\end{aligned}$} & {$\!\begin{aligned}&(L||R)\leftarrow FE(L,R)\\&(L||R)\leftarrow FLL(L,R,kl_0,kl_1)\end{aligned}$} \\
        \midrule
        6-10 & $(L||R)\leftarrow FE(L,R,k_r)$ & $(L||R)\leftarrow FE(L,R,k_r)$ \\
        \midrule
        11 & {$\!\begin{aligned}&(L||R)\leftarrow FE(L,R)\\&(L||R)\leftarrow FLL(L,R,kl_2,kl_3)\end{aligned}$} & {$\!\begin{aligned}&(L||R)\leftarrow FE(L,R)\\&(L||R)\leftarrow FLL(L,R,kl_2,kl_3)\end{aligned}$} \\
        \midrule
        12-16 & $(L||R)\leftarrow FE(L,R,k_r)$ & $(L||R)\leftarrow FE(L,R,k_r)$ \\
        \midrule
        17 & & {$\!\begin{aligned}&(L||R)\leftarrow FE(L,R)\\&(L||R)\leftarrow FLL(L,R,kl_4,kl_5)\end{aligned}$} \\
        \midrule
        18-23 & & $(L||R)\leftarrow FE(L,R,k_r)$ \\
        \bottomrule
    \end{tabular}
\end{table}

@@

\[
    c=(R||L)\oplus (kw_2)||kw_3)
\]

@@

We will give an overview of the main functional components of Camellia.

\paragraph{Feistel round function $FE$}

\begin{align*}
    FE:(\F{64})^3 &\rightarrow (\F{64})^2 \\
    (L_{(64)},R_{(64)},k_{(64)})&\mapsto (R\oplus F(L,k), L)
\end{align*}

\paragraph{SP-function $F$}

\begin{align*}
    F:(\F{64})^2 &\rightarrow \F{64} \\
    (X_{(64)},k_{(64)})&\mapsto P(S(X\oplus k))
\end{align*}

\paragraph{Substitution function $F$}

\begin{align*}
    S:\F{64} &\rightarrow \F{64} \\
    l_{0(8)||l_{1(8)||l_{2(8)||l_{3(8)||l_{4(8)||l_{5(8)||l_{6(8)||l_{7(8)||&\mapsto P(S(X\oplus k))
\end{align*}

\paragraph{$FL$ layer function $FLL$}

\begin{align*}
    FLL:(\F{64})^4 &\rightarrow (\F{64})^2 \\
    (L_{(64)},R_{(64)},k_{0(64)},k_{1(64)})&\mapsto (FL(L,k_0),FL^{-1}(R,k_1))
\end{align*}

@@

\begin{table}[h!]
    \centering
    \caption{ARMv8 profiles}
    \begin{tabularx}{\textwidth}{lX}
        \toprule
        Profile & Description \\
        \midrule
        Application (A) & Traditional use with virtual memory and privilege level support \\
        Real-time (R) & Real-time, low-latency, deterministic embedded systems \\
        Microcontroller (M) & Very low-power, fast-interrupt embedded systems \\
        \bottomrule
    \end{tabularx}
\end{table}

\begin{table}[h!]
    \centering
    \caption{ARMv8 execution states}
    \begin{tabularx}{\textwidth}{llX}
        \toprule
        Execution state & Usage & Instruction sets \\
        \midrule
        AArch32 & 32-bit compatibility & A32/T32 \\
        AArch64 & 64-bit & A64 \\
        \bottomrule
    \end{tabularx}
\end{table}

@@

\begin{table}[h!]
    \centering
    \small
    \caption{A64 addressing modes}
    \begin{tabularx}{\textwidth}{llX}
        \toprule
        Addressing mode & Example & Description \\
        \midrule
        Base register & \texttt{LDR W0, [X1]} & \texttt{W0 = *(X1);} \\
        Offset & \texttt{LDR W0, [X1, \#12]} & \texttt{W0 = *(X1 + 12);} \\
        Pre-indexing & \texttt{LDR W0, [X1, \#12]!} & \texttt{X1 += 12; W0 = *(X1);} \\
        Post-indexing & \texttt{LDR W0, [X1], \#12} & \texttt{W0 = *(X1); X1 += 12;} \\
        \bottomrule
    \end{tabularx}
\end{table}

@@

\begin{table}[h!]
    \centering
    \footnotesize
    \caption{Common NEON intrinsics}
    \begin{tabularx}{\textwidth}{llX}
        \toprule
        Intrinsic && Summary \\
        \midrule
        \texttt{uint64\_t} & \texttt{vgetq\_lane\_u64(void)} & Extract a single lane \\
        \midrule
        \texttt{void} & \texttt{vsetq\_lane\_u64(uint64\_t)} & Insert a single lane \\
        \midrule
        \texttt{uint64x2\_t} & \texttt{vdupq\_n\_u64(uint64\_t)} & Initialize all lanes to same value \\
        \midrule
        \texttt{void} & \texttt{vst1q\_u64(uint64\_t*, uint64x2\_t)} & Store from register to memory \\
        \midrule
        \texttt{uint64x2\_t} & \texttt{vld1q\_u64(uint64\_t*, uint64x2\_t)} & Load from memory to register \\
        \midrule
        \texttt{uint8x16\_t} & \texttt{veorq\_u8(uint8x16\_t, uint8x16\_t)} & bitwise XOR \\
        \midrule
        \texttt{uint8x16\_t} & \texttt{vandq\_u8(uint8x16\_t, uint8x16\_t)} & bitwise AND \\
        \midrule
        \texttt{uint8x16\_t} & \texttt{vorrq\_u8(uint8x16\_t, uint8x16\_t)} & bitwise OR \\
        \midrule
        \texttt{uint8x16\_t} & \texttt{vmvnq\_u8(uint8x16\_t)} & bitwise NOT \\
        \midrule
        \texttt{uint8x16\_t} & \texttt{vqtbl2q\_u8(uint8x16\_t, uint8x16\_t)} & permutation (\texttt{TBL}) \\
        \bottomrule
    \end{tabularx}
\end{table}
